A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-07 podc_paper/algorithm-2.tex
Updated code to make GC run node-by-node instead

@@

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
	\gc{\State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}}+1,$\\
		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}},$\\
		$\fld{index} \assign\ h+1$}}\label{enqNewGC}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
	\gc{\State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}},$\\
	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}+1,$\\
	    $\fld{index}\assign h+1$}}\label{deqNewGC}
    \State \Call{Append}{\var{B}}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \gc{$h+1$}, 1}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}
    \label{deqRestGC}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{Insert}{\var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
    \gc{\State \here{Need to clean out my own leaf sometimes--how to determine what threshold to use for the split?}}
\EndFunction{Append}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{\gc{GCRefresh}}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
        \State \Call{\gc{GCRefresh}}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

	\gc{
\Function{boolean}{GCRefresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}, with GC}
	\State $\var{T} \assign\ \var{v}.\fld{blocks}$\label{readTGC}
    \State $\var{h} \assign\ \Call{MaxBlock}{T}.\fld{index}$ \label{readHeadGC}
	\If{\var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
		\State \linecomment{garbage collection phase}
		\State \var{max} \assign 0
		\For {$k \assign 1..p$}
			\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
		\EndFor
		\State \Call{Help}{}
		\State \Return $\Call{Discard}{\var{root}, \var{max}-1}$\label{discard}
    \Else\ \linecomment{do a standard refresh}
        \State $T' \assign \Call{RefreshUpdate}{T, v}$\label{RU1}
        \If{$T' = \nl$} \Return{\tr}\label{emptyblockGC}
    	\Else\ \Return $\Call{CAS}{\var{v}.\fld{blocks}, T, T'}$\label{casGC}
		\EndIf
	\EndIf
%    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
\EndFunction{GCRefresh}
}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{48}
\gc{\Function{RBT}{RefreshUpdate}{\typ{RBT} \var{T}, \typ{Node} \var{v}}
    \State \linecomment{append new block to RBT $T$ read from $v.\fld{blocks}$}
    \State \Call{SetSuper}{$v$, $T$}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
    \Else \ \Return \Call{Insert}{$T$, \var{new}}\label{refreshInsertGC}
    \EndIf
\EndFunction{RefreshUpdate}
}
\spac

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \gc{
    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
    \State \var{new}.\fld{index} \assign \var{i}
    }
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}\label{CBnullGC}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}
\end{algorithmic}
\end{minipage}
\caption{Bounded-space queue implementation.\label{pseudocode1GC}}
\end{figure}

@@

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{last} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
    	\State \Call{Propagate}{$\ell.\fld{parent}$}
        \If{$\var{last}.\fld{element}\hspace{-.1mm} =\hspace{-.1mm} \nl$\hspace{-.15mm} and\hspace{-.15mm} $\var{last}.\fld{index}\hspace{-.12mm}>\hspace{-.12mm}0$} 
            \State \linecomment{operation is a  dequeue}
            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, \var{last}.\fld{index}, 1}$
            \State $\var{last}.\fld{response} \assign\ \Call{FindResponse}{b, i}$ 
        \EndIf
    \EndFor
\EndFunction{Help}

\Function{boolean}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
    \State \linecomment{discard blocks with index less than or equal}
    \State \linecomment{to $t$ from $v$'s RBT and their subblocks}
    \If{$v$ is an internal node}
	    \State \var{t\sub{left}} \assign \var{v}.\fld{blocks}[\var{t}].\eleft
    	\State \var{t\sub{right}} \assign \var{v}.\fld{blocks}[\var{t}].\eright
		\For{\var{j} \assign 1 to $p$}
        	\State $T \assign \var{v}.\fld{blocks}$\label{readT2GC}
        	\State $T'  \assign \Call{Split}{T,t}$\label{split}
			\State \linecomment{\op{Split} removes blocks with $\var{index} < t$}
			\State $T'' \assign \Call{RefreshUpdate}{T',v}$ \label{RU2}
        	\State \hangbox{$\var{result} \assign \Call{CAS}{\var{v}.\fld{blocks}, T, \\
					\hspace*{10mm}(T'' = \nl\ ?\ T' : T'')$}}\label{splitCASGC}
        	\State \bf{exit when} \var{result} 
	        \State \linecomment{if \op{CAS} succeeds go to line \ref{discardrecursecomment}}
    	\EndFor
    	\If{\var{v} is an internal \typ{Node}} 
			\State \linecomment{discard subblocks in descendants of $v$}\label{discardrecursecomment}
        	\State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}\label{discardrecurse1}
        	\State \Call{Discard}{\var{v}.\fld{right}, \var{t\sub{right}}}\label{discardrecurse2}
    	\EndIf
	\EndIf
	\State \Return \var{result}
\EndFunction{Discard}
}
\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] has been propagated to root}
    \State \linecomment{and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
    \Else
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$
	    \EndIf
	    \State \linecomment{compute index \var{i} of dequeue in superblock}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
        \EndIf \label{computeISuperEndGC}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
        \gc{\State \var{last}[\var{this process id}] \assign\ \var{b}}\label{setlast1}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
        \gc{
        \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
        \State \var{last}[\var{this process id}] \assign\ $b_e$\label{setlast2}
        \State \Return{\var{e}}
        }
    \EndIf
\EndFunction{FindResponse}

@@

\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}

    \If{\var{v} is a leaf node}
        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRightGC} \cmt{required enqueue is in \var{v}.\fld{left}}
            \State \fld{dir} \assign\ \fld{left}
        \Else
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        \EndIf
        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t. $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChildGC}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$}
    \EndIf
\EndFunction{GetEnqueue}

>>>>

\HERE{Idea:  replace last array with a bit in blocks that gets set when somebody dequeues an element in the block.  then can use the max index of any marked block minus 1 as the last completed block.  
In order to ensure that blocks of dequeues are also cleared out,
maybe dequeues that return null have to mark all blocks along the dequeue's path?
}

@@

\here{Emphasize that RBT nodes have \emph{pointers to} blocks.  Thus when RBT does node copying,
it copies the pointer to the block; it does not make a copy of the block itself.
This is important because if we mark a block that has been node-copied, we want the block to still be marked.  Similarly for super field.}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \gc{
    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
	\State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}}+1,$\\
		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}$,\\
		$\fld{index}\assign h+1$}\label{enqNewGC}
	}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \gc{
    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}},$\\
	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}+1$,\\
	    $\fld{index}\assign h+1$}\label{deqNewGC}
	}
    \State \Call{Append}{\var{B}}
    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, \gc{h+1}, 1, \fa}$
    \gc{
    \State $\var{res} \assign \Call{FindResponse}{b, i}$
	\If{$\var{res} = \nl$} \linecomment{mark blocks}
		\State $\Call{IndexDequeue}{\var{leaf}, h+1, 1, \tr}$
	\EndIf
	\State \Return{\var{res}}
    }
    \label{deqRestGC}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

\gc{
\Function{RBT}{AddBlock}{\typ{RBT} \var{T}, \typ{Block} \var{B}}
    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
    \If{\var{B}.\fld{index} is a multiple of $p^2$}
    	\State \linecomment{Do garbage collection}
        \State \Call{Help}{}
        \State $m$ \assign\ maximum \fld{index} of marked block in $T$
        \State $T' \assign\ \Call{Split}{T,m-1}$
        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
        \State \Return \Call{Insert}{$T',B$}
	\Else\ \Return \Call{Insert}{$T, B$}
    \EndIf
\EndFunction{AddBlock}
}
\spac

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
	\gc{
	\State $T \assign v.\fld{blocks}$
    \State \Call{SetSuper}{$v$, $T$}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
    \Else  
    	\gc{
	    \State $T' \assign \Call{AddBlock}{T, \var{new}}$
		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}
		}
    \EndIf

\EndFunction{Refresh}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \gc{
    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
    \State $\var{new}.\fld{index} \assign i$
    }
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}\label{CBnullGC}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
    \EndIf
\EndFunction{FindResponse}

\end{algorithmic}
\end{minipage}
\caption{Bounded-space queue implementation.\label{pseudocode1GC}}
\end{figure}

@@

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{last} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
    	\State \Call{Propagate}{$\ell.\fld{parent}$}
        \If{$\var{last}.\fld{element}\hspace{-.1mm} =\hspace{-.1mm} \nl$\hspace{-.15mm} and\hspace{-.15mm} $\var{last}.\fld{index}\hspace{-.12mm}>\hspace{-.12mm}0$} 
            \State \linecomment{operation is a  dequeue}
            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, \var{last}.\fld{index}, 1}$
            \State $\var{last}.\fld{response} \assign\ \Call{FindResponse}{b, i}$ 
        \EndIf
    \EndFor
\EndFunction{Help}
}
\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}, \gc{\typ{boolean} \var{mark}}}
    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] has been propagated to root}
    \State \linecomment{and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
    \gc{
    \If{\var{mark}} \var{v}.\fld{blocks}[\var{b}].\fld{mark} \assign \tr
    \EndIf
    }
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
    \Else
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$
	    \EndIf
	    \State \linecomment{compute index \var{i} of dequeue in superblock}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
        \EndIf \label{computeISuperEndGC}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}
	\gc{\State $\var{v}.\fld{blocks}[\var{b}].\fld{marked} := \tr$}
    \If{\var{v} is a leaf node}
        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRightGC} \cmt{required enqueue is in \var{v}.\fld{left}}
            \State \fld{dir} \assign\ \fld{left}
        \Else
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        \EndIf
        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t. $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChildGC}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$}
    \EndIf
\EndFunction{GetEnqueue}
